OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-11-24 manuscript_nar.tex
Adding supp tables in excel with legend + last changes

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 5~kbp.
%\textit{OneD} is implemented as an R package
%available at http://www.github.com/qenvio/dryhic.
\end{abstract}

@@

Here we propose \textit{OneD}, a method to correct local chromosomal
abnormalities in Hi-C experiments. \textit{OneD} explicitly models the
contribution of known biases via a generalized additive model. The
normalized data is more reproducible between replicates and across
different protocols. Importantly, \textit{OneD} is also efficient when
cells have a normal karyotype, where it performs as well as the best
normalization methods. Finally, the implementation is as fast as
\textit{ICE} and it scales up to 5~kbp resolution with reasonable
computing resources.

@@

\noindent
where $x_{ij}$ is the raw number of contacts between bins $i$ and $j$, and
$z_k$ is the additive bias of genomic feature $k$. The smooth functions
$\{f_k(\cdot)\}$ are estimated jointly with the negative binomial
dispersion parameter $\theta$ using the \texttt{mgcv} package
\citep{wood2011fast} of the R software \citep{coreteam2014r}.

@@

In line with previous methods
\citep{yaffe2011probabilistic,hu2012hicnorm}, the default features
used to fit the model are the local G+C content, the read mappability and
the content and number of restriction sites. The model and the
implementation can be modifed or extended with any user-provided genomic
features.

@@

We developed the routines contained in the \texttt{dryhic} R package
(available at http://www.github.com/qenvio/dryhic) to efficiently create
sparse representations of contact matrices and further apply
\textit{vanilla}, \textit{ICE}, \textit{KR} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. All the results are based on a
resolution of 100~kbp, but we found no major differences at different
resolutions (not shown).

@@

Simulations were based on altering four experimental Hi-C samples obtained
from diploid mouse cell lines (either B cell or ES cell, and from either
CRG or UCSC). The simulation strategy was the following: First we selected
uniformly at random the amount of copy number break points (from 3 to 10).
We then placed the break points along chromosomes 18 and 19 uniformly at
random. Next, we assigned a copy number (2, 3, 4 or 10 with equal
probability) to each segment delimited by the break points. We computed
the outer product of this simulated copy number profile and multiplied it
element-wise with the original contact matrices of chromosomes 18 and 19.
The resulting matrices were used as input for correction methods. For each
simulation (100 in total), we measured the pairwise reproducibility score
defined by \citet{yan2017hicspector} and computed the average for pairs
from the same cell type minus the average for pairs from different cell
types.

@@

To measure the gain or loss of similarity upon normalization, we compared
raw matrices to obtain a baseline. The differences with this baseline were
estimated using a linear mixed model fitted with the \texttt{lmer}
function of the \texttt{lme4} R package \citep{bates2015lme4}, where the
fixed effect was the normalization method and the random effect was the
chromosome. Receiver operating characteristic (ROC) curves were computed
using the \texttt{ROCR} package \citep{sing2005rocr} for the three first sets
(those including samples of different cell types).

@@

Optionally, \textit{OneD} allows the user to also correct the Hi-C signal
for the copy number. In this case, a hidden Markov model is fitted on the
corrected contact profile in order to infer the local copy number. Each
cell of the Hi-C matrix is then further corrected using equation
(\ref{eq:cnvnorm}), \textit{i.e.} it is divided by the square root of the
inferred copy number. In essence, this approach flatens the contact
profile so that it is no longer proportional to the local copy number. The
light blue line in figure~\ref{fig:totals}B shows the result of this
process. After this correction, the contact profile fluctuates around a
genome-wide constant value.

@@

We collected multiple Hi-C data sets obtained from T47D and K562 cells,
both with aberrant karyotypes. The first pool, referred to as the T47D data
set contained six T47D samples and two K562 samples, the second, refered
to as the K562 data set contained eight K562 samples and two T47D samples.
We compared matrices before and after normalization by different methods
(see Materials and Methods, distributions shown in Supplementary Figures~2
and 3). This gave an indication of the impact of a given normalization
method on the biological variation (differences between samples from T47D
and K562) and on the technical variation (differences among samples from
the same cell type). The results are summarized in
Figure~\ref{fig:aberrant}.

@@

Does the performance of \textit{OneD} on aberrant karyotypes come at the
cost of decreased performance on normal karyotypes? To address this
question, we collected data from mouse B cells and embryonic stem (ES)
cells, both with a normal karyotype. The cell types were pooled in almost
equal proportions (see Supplementary Table~1) and we performe the same
tests as above using the same metrics (distributions shown in
Supplementary Figure~7)

In this test, the reproducibility between different cell types was barely
affected (Figure~\ref{fig:normal}A). All the methods increased the
reproducibility between identical cell types (except one, see below), but
only when the experimental conditions were different. This means that all
the methods were able to remove some technical biases. \textit{ICE} and
\textit{KR} showed the stongest increase in reproducibility, but
\textit{OneD}, \textit{OneD+CN}, and \textit{caICB} had competitive
performance. \textit{vanilla} and \textit{LGF} were less competitive, and
\textit{raw+CN} (copy number correction alone) performed poorly, because
it is equivalent to no normalization on diploid cells. Other metrics gave
similar results (Supplementary Figures~8, 9 and 10).

As above, we used the reproducibility score for classification and
compared the tools with a ROC curve (Figure~\ref{fig:normal}B). The
performance of all the tools were high and less variable than for
aneuploid cell lines. \textit{OneD} achieved the highest area under the
curve on this problem, but with a small margin over \textit{KR} and
\textit{ICE}. The performance of \textit{OneD+CN} was the same as that of
\textit{OneD} because copy numer correction has no effect on diploid cell
lines. Using other metrics to compare matrices gave similar results
(Supplementary Figures~8, 9 and 10).

We took advantage of the wide range of protocols used in the GM12878 set to
further investigate the performance of each method. The differences between
methods were minor and \textit{OneD} showed the greatest improvement in
reporducibility (Supplementary Figure~11).

@@

To test this idea, we collected a Hi-C data set from breast cancer cell
lines and measured the reproducibility score bewteen identical cell types
after normalization with the different tools. In this case, we found that
\textit{OneD+CN} outperformed \textit{OneD} and gave the highest overall
reproducibility (Figure~\ref{fig:cnv_correction}A). Even though the
variability is high in this case, this suggests that the karyotypes of
these cells have diverged, so that the copy number correction performed by
\textit{OneD} increased the reproducibility between samples from the same
cell type.

@@

@@

We reasoned that \textit{OneD} should handle such cases gracefully,
because the signal is formally equivalent to an aberrant karyotype where
most of the genome is present in zero copy. We thus evaluated the
performance of \textit{OneD} on a Capture Hi-C data set at 5~kbp
resolution centered on a 6~Mbp domain of chromosome 6 in T47D cells. We
found that the reproducibility score between replicates was higher after
normalization with \textit{OneD} than with matrix-balancing methods
(figure~\ref{fig:chic}A). The performance of \textit{OneD} was only
slightly lower than the raw data, suggesting that it barely disturbs the
biological signal. Other tools were not included beause they either
crashed (\textit{caICB}) or had prohibitive run time (\textit{LGF}).

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_5.eps}}
\caption{Bias removal on Capture Hi-C data A. Performance on the Capture
Hi-C data. Two replicates centered on a 6~Mbp domain on chromosome 6 in
T47D cells were normalized by the indicated tools and their
reproducibility score was computed. B. Performance on virtual Capture Hi-C
data. Two Hi-C experiments were post-processed to produce an output
similar to the Capture Hi-C experiment depicted in panel A. The signal
outside the domain on chromosome 6 was removed and the data were
normalized (capt), or the data were normalized before removing the signal
oustside the domain on chromosome 6 (full). Four matrices per method were
generated and the reproducibility score between the pairs from different
replicates were computed. Note that the scores are identical for the raw
data because in this case the matrices of the same replicate are
identical.}
\label{fig:chic}
\end{figure}

To gain additional insight, we also set up a ``virtual Capture Hi-C'' data
set by removing the signal outside the same region of chromosome 6 from
Hi-C experiments performed in the same cell type. In this setup, we could
restrict the data and then normalize, or normalize and then restrict the
data. This allowed us to probe how sensitive the methods are to the shape
of the input data. Here we found that \textit{OneD} significantly
outperformed the other methods: it featured the highest reproducibility
score between replicates with practically no influence from the shape of
the data (figure~\ref{fig:chic}B). Also, in this case, \textit{OneD}
enhanced the reproducibility bewteen replicates compared to the raw data.
This confirms the view that capture methods are a natural framework for
the statistical model underlying \textit{OneD}. In comparison,
matrix-balancing methods showed lower reproducibility between replicates,
together with more variability due to the shape of the data.

@@

\textit{OneD} corrects a single variable instead of the whole matrix, and
thus estimates the model parameters much faster than previous explicit
methods. We measured the running time of the different tools on a 3.3~GHz
machine with 62 GB~RAM, always using the default parameters.
Figure~\ref{fig:times}A shows the running times of the different methods
on all the samples shown in Supplementary Figure~1 at 100~kbp resolution.
The fastest method was \textit{vanilla} and the slowest \textit{LGF}, with
an over 100-fold span between the two.

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_6.eps}}
\caption{Computing time of the bias correction methods. A. Total time to
process samples listed in Supplementary Table~1 at 100~kbp resolution.
Each dot corresponds to a sample. B. Total time to process sample HIC070
at the indicated bin size and sequencing depths. Note the logarithmic
scale on the y-axis on both panels.}
\label{fig:times}
\end{figure}

To better understand these results, we also investigated the influence of
bin size and sequencing depth on the running time. We performed a
benchmark on the sample with highest sequencing depth in the data set
(HIC070, see Supplementary Table~1) at different bin sizes while
downsampling the contact matrices to mimic lower resolutions. On this
test, we included only the methods competing with \textit{OneD} in terms
of speed, \textit{i.e.} the matrix-balacing methods.
Figure~\ref{fig:times}B shows that the running time of \textit{OneD} does
not depend on the sequencing depth. Strikingly, \textit{OneD} also gives
nearly identical results at very different sequencing depths
(Supplementary Figure~13). \textit{vanilla} was the overall fastest method
and \textit{OneD} was usually faster than \textit{ICE} and \textit{KR}
at high but not at low sequencing depth. At low resolution (100~kbp) the
speed advantage of \textit{OneD} appeared from low sequencing depth, but
at high resolution reslution (5~kbp) it appeared only at high sequening
depth.

>>>>

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 5~Kbp.
%\textit{OneD} is implemented as an R package
%available at http://www.github.com/qenvio/dryhic.
\end{abstract}

@@

Here we propose \textit{OneD}, a method to correct local chromosomal
abnormalities in Hi-C experiments. \textit{OneD} explicitly models the
contribution of known biases via a generalized additive model. The
normalized data is more reproducible between replicates and across
different protocols. Importantly, \textit{OneD} is also efficient when
cells have a normal karyotype, where it performs as well as the best
normalization methods. Finally, the implementation is as fast as
\textit{ICE} and it scales up to 5~Kbp resolution with reasonable
computing resources.

@@

\noindent
where $x_{ij}$ is the raw number of contacts between bins $i$ and $j$, and
$z_k$ is the additive bias of genomic feature $k$. The smooth functions
$\{f_k(\cdot)\}$ are estimated jointly with the negative binomial
dispersion parameter $\theta$ using the the default arguments of the
\texttt{mgcv::s} and \texttt{mgcv::nb} functions \citep{wood2011fast} of the R
software \citep{coreteam2014r}.

@@

In line with previous methods
\citep{yaffe2011probabilistic,hu2012hicnorm}, the default features
used to fit the model are the local G+C content, the read mappability and
the number of restriction sites. The model and the
implementation can be modified or extended with any user-provided genomic
features.

@@

We developed the routines contained in the \texttt{dryhic} R package
(available at http://www.github.com/qenvio/dryhic) to efficiently create
sparse representations of contact matrices and further apply
\textit{vanilla}, \textit{ICE}, \textit{KR} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. All the results are based on a
resolution of 100~Kbp, but we found no major differences at different
resolutions (not shown).

@@

Simulations were based on altering four experimental Hi-C samples obtained
from diploid mouse cell lines (b7fa2d8db\_bfac48760, fc3e8b36a\_7bf1bf374,
GSM987817 and GSM862720), either B cell or ES cell, and from either
CRG or UCSC. The simulation strategy was the following: First we selected
uniformly at random the amount of copy number break points (from 3 to 10).
We then placed the break points along chromosomes 18 and 19 uniformly at
random. Next, we assigned a copy number (2, 3, 4 or 10 with equal
probability) to each segment delimited by the break points. We computed
the outer product of this simulated copy number profile and multiplied it
element-wise with the original contact matrices of chromosomes 18 and 19.
The resulting matrices were used as input for correction methods. For each
simulation (100 in total), we measured the pairwise reproducibility score
defined by \citet{yan2017hicspector} and computed the average for pairs
from the same cell type minus the average for pairs from different cell
types.

@@

To measure the gain or loss of similarity upon normalization, we compared
raw matrices to obtain a baseline. The differences with this baseline were
estimated using a linear mixed model fitted with the \texttt{lmer}
function of the \texttt{lme4} R package \citep{bates2015lme4}, where the
fixed effect was the normalization method and the random effect was the
chromosome. To test the ability of the different methods to separate samples
from different cell origin we generated receiver operating characteristic (ROC) curves
using the similarity metric (e.g. reproducibility score) as the classifier score
and the relative cell type (i.e. same or different cell type) as the binary
classification. ROC curves were computed using the \texttt{ROCR} package
\citep{sing2005rocr} for the three first sets (those including samples of
different cell types).

@@

Optionally, \textit{OneD} allows the user to also correct the Hi-C signal
for the copy number. In this case, a hidden Markov model is fitted on the
corrected contact profile in order to infer the local copy number. Each
cell of the Hi-C matrix is then further corrected using equation
(\ref{eq:cnvnorm}), \textit{i.e.} it is divided by the square root of the
inferred copy number. In essence, this approach flattens the contact
profile so that it is no longer proportional to the local copy number. The
light blue line in figure~\ref{fig:totals}B shows the result of this
process. After this correction, the contact profile fluctuates around a
genome-wide constant value.

@@

We collected multiple Hi-C data sets obtained from T47D and K562 cells,
both with aberrant karyotypes. The first pool, referred to as the T47D data
set contained six T47D samples and two K562 samples, the second, referred
to as the K562 data set contained eight K562 samples and two T47D samples.
We compared matrices before and after normalization by different methods
(see Materials and Methods, distributions shown in Supplementary Figures~2
and 3). This gave an indication of the impact of a given normalization
method on the biological variation (differences between samples from T47D
and K562) and on the technical variation (differences among samples from
the same cell type). The results are summarized in
Figure~\ref{fig:aberrant} and Supplementary Table~3.

@@

Does the performance of \textit{OneD} on aberrant karyotypes come at the
cost of decreased performance on normal karyotypes? To address this
question, we collected data from mouse B cells and embryonic stem (ES)
cells, both with a normal karyotype. The cell types were pooled in almost
equal proportions (see Supplementary Table~1) and we performed the same
tests as above using the same metrics (distributions shown in
Supplementary Figure~7)

In this test, the reproducibility between different cell types was barely
affected (Figure~\ref{fig:normal}A). All the methods increased the
reproducibility between identical cell types (except one, see below), but
only when the experimental conditions were different. This means that all
the methods were able to remove some technical biases. \textit{ICE} and
\textit{KR} showed the strongest increase in reproducibility, but
\textit{OneD}, \textit{OneD+CN}, and \textit{caICB} had competitive
performance. \textit{vanilla} and \textit{LGF} were less competitive, and
\textit{raw+CN} (copy number correction alone) performed poorly, because
it is equivalent to no normalization on diploid cells. Other metrics gave
similar results (Supplementary Figures~8, 9 and 10, Supplementary Table~3).

As above, we used the reproducibility score for classification and
compared the tools with a ROC curve (Figure~\ref{fig:normal}B). The
performance of all the tools were high and less variable than for
aneuploid cell lines. \textit{OneD} achieved the highest area under the
curve on this problem, but with a small margin over \textit{KR} and
\textit{ICE}. The performance of \textit{OneD+CN} was the same as that of
\textit{OneD} because copy number correction has no effect on diploid cell
lines. Using other metrics to compare matrices gave similar results
(Supplementary Figures~8, 9 and 10).

We took advantage of the wide range of protocols used in the GM12878 set to
further investigate the performance of each method. The differences between
methods were minor and \textit{OneD} showed the greatest improvement in
reproducibility (Supplementary Figure~11).

@@

To test this idea, we collected a Hi-C data set from breast cancer cell
lines and measured the reproducibility score between identical cell types
after normalization with the different tools. In this case, we found that
\textit{OneD+CN} outperformed \textit{OneD} and gave the highest overall
reproducibility (Figure~\ref{fig:cnv_correction}A). Even though the
variability is high in this case, this suggests that the karyotypes of
these cells have diverged, so that the copy number correction performed by
\textit{OneD} increased the reproducibility between samples from the same
cell type.

@@

Besides matrix correction, copy number estimation could have a potential stand-alone
utility. For instance, we applied \textit{OneD+CN} to a T47D sample
(dc3a1e069\_51720e9cf) at 10 Kbp and compared the correlation with the
array-based copy number. The correlation increased from .80 for the raw totals,
to .88 for the \textit{OneD} corrected totals to .92 for the estimated CN. We
provide the table of the estimated start and end points as Supplementary Table~
2.

@@

We reasoned that \textit{OneD} should handle such cases gracefully,
because the signal is formally equivalent to an aberrant karyotype where
most of the genome is present in zero copy. We thus evaluated the
performance of \textit{OneD} on a Capture Hi-C data set at 5~Kbp
resolution centered on a 6~Mbp domain of chromosome 6 in T47D cells. We
found that the reproducibility score between replicates was higher after
normalization with \textit{OneD} than with matrix-balancing methods
(figure~\ref{fig:chic}A). The performance of \textit{OneD} was only
slightly lower than the raw data, suggesting that it barely disturbs the
biological signal. Other tools were not included beaus they either
crashed (\textit{caICB}) or had prohibitive run time (\textit{LGF}).

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_5.eps}}
\caption{Bias removal on Capture Hi-C data A. Performance on the Capture
Hi-C data. Two replicates centered on a 6~Mbp domain on chromosome 6 in
T47D cells were normalized by the indicated tools and their
reproducibility score was computed. B. Performance on virtual Capture Hi-C
data. Two Hi-C experiments were post-processed to produce an output
similar to the Capture Hi-C experiment depicted in panel A. The signal
outside the domain on chromosome 6 was removed and the data were
normalized (capt), or the data were normalized before removing the signal
outside the domain on chromosome 6 (full). Four matrices per method were
generated and the reproducibility score between the pairs from different
replicates were computed. Note that the scores are identical for the raw
data because in this case the matrices of the same replicate are
identical.}
\label{fig:chic}
\end{figure}

To gain additional insight, we also set up a ``virtual Capture Hi-C'' data
set by removing the signal outside the same region of chromosome 6 from
Hi-C experiments performed in the same cell type. In this setup, we could
restrict the data and then normalize, or normalize and then restrict the
data. This allowed us to probe how sensitive the methods are to the shape
of the input data. Here we found that \textit{OneD} significantly
outperformed the other methods: it featured the highest reproducibility
score between replicates with practically no influence from the shape of
the data (figure~\ref{fig:chic}B). Also, in this case, \textit{OneD}
enhanced the reproducibility between replicates compared to the raw data.
This confirms the view that capture methods are a natural framework for
the statistical model underlying \textit{OneD}. In comparison,
matrix-balancing methods showed lower reproducibility between replicates,
together with more variability due to the shape of the data.

@@

\textit{OneD} corrects a single variable instead of the whole matrix, and
thus estimates the model parameters much faster than previous explicit
methods. We measured the running time of the different tools on a 3.3~GHz
machine with 62 GB~RAM, always using the default parameters.
Figure~\ref{fig:times}A shows the running times of the different methods
on all the samples shown in Supplementary Figure~1 at 100~Kbp resolution.
The fastest method was \textit{vanilla} and the slowest \textit{LGF}, with
an over 100-fold span between the two.

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {nar_figures/figure_6.eps}}
\caption{Computing time of the bias correction methods. A. Total time to
process samples listed in Supplementary Table~1 at 100~Kbp resolution.
Each dot corresponds to a sample. B. Total time to process sample HIC070
at the indicated bin size and sequencing depths. Note the logarithmic
scale on the y-axis on both panels.}
\label{fig:times}
\end{figure}

To better understand these results, we also investigated the influence of
bin size and sequencing depth on the running time. We performed a
benchmark on the sample with highest sequencing depth in the data set
(HIC070, see Supplementary Table~1) at different bin sizes while
downsampling the contact matrices to mimic lower resolutions. On this
test, we included only the methods competing with \textit{OneD} in terms
of speed, \textit{i.e.} the matrix-balancing methods.
Figure~\ref{fig:times}B shows that the running time of \textit{OneD} does
not depend on the sequencing depth. Strikingly, \textit{OneD} also gives
nearly identical results at very different sequencing depths
(Supplementary Figure~13). \textit{vanilla} was the overall fastest method
and \textit{OneD} was usually faster than \textit{ICE} and \textit{KR}
at high but not at low sequencing depth. At low resolution (100~Kbp) the
speed advantage of \textit{OneD} appeared from low sequencing depth, but
at high resolution resolution (5~Kbp) it appeared only at high sequencing
depth.
